---
title: Why You Shouldn't Use <br>parallel::detectCores() in R
author: StatistikinDD
subtitle: Created `r format(Sys.Date(), format = "%Y-%m-%d")`
format: 
  revealjs:
    theme: simple
    logo: StatistikinDD.png
    highlight-style: github
execute:
  echo: true
  eval: false
editor: visual
---

## Parallelization: `detectCores()`

-   `detectCores()` from Base R's `parallel` package is very popular

-   ***Don't use it!!***

. . .

-   Short: Use `parallelly::availableCores()` instead

-   Long version: Why not?

. . .

-   Which problems may arise if I do?

## Credit to *Henrik Bengtsson*

Blog post:\
***Please Avoid `detectCores()` in your R Packages***\
[https://www.jottr.org/2022/12/05/avoid-detectcores/](https://www.jottr.org/2022/12/05/avoid-detectcores/) <br><br>

| Profile | Link                                          |
|---------|-----------------------------------------------|
| Website | [https://jottr.org](https://jottr.org) <br>[https://futureverse.org](https://futureverse.org) |
| Github  | [https://github.com/HenrikBengtsson](https://github.com/HenrikBengtsson)            |
| Twitter | @henrikbengtsson                              |

## Credit to *Henrik Bengtsson* {.smaller}

| Package      | Description                                                         |
|-----------------------------|-------------------------------------------|
| future       | Unified Parallel and Distributed Processing in R for Everyone       |
| future.apply | Apply Function to Elements in Parallel using Futures                |
| progressr    | An Inclusive, Unifying API for Progress Updates                     |
| matrixStats  | Methods that Apply to Rows and Columns of Matrices (and to Vectors) |
| parallelly   | Enhancing the parallel Package                                      |

... and many more!


## Problems with `detectCores()`

(@) `detectCores()` may return a missing value

(@) `detectCores()` may return one

(@) `detectCores()` may return too many cores

(@) `detectCores()` does not give the number of "allowed" cores

::: {.notes}
  + on a personal computer
  + on a shared computer
  + on a shared Compute Cluster
  + in a Linux container using CGroups
:::


## 1. `detectCores()` may return a missing value 

see `help("detectCores", package = "parallel")`  
Value: An integer, ***NA if the answer is unknown***

Example:

. . . 

```{r}
ncores <- detectCores()
workers <- parallel::makeCluster(ncores)
```

<div style="color: red">
Error in makePSOCKcluster(names = spec, ...) : 
numeric 'names' must be >= 1
</div>

. . . 

Solution:
```{r}
#| code-line-numbers: "1"
parallelly::availableCores()
```


## 2. `detectCores()` may return one

<br>
So this common usage pattern will fail:

<br>
```{r}
ncores <- detectCores() - 1L  # may return 0
```

. . . 

<br>
Solution:

```{r}
parallelly::availableCores()  # min = 1
parallelly::availableCores(omit = 1)  # leave one core, if possible; min = 1
```

## 3. `detectCores()` may return too many cores

* R has a limit in the number of connections it can have open at any time
* For R 4.2.2, the *theoretical* limit is **125 open connections**
* The *practical* limit is **lower**: Connections may be in use elsewhere


## 3. `detectCores()` may return too many cores {.smaller}

Examples for a 192-core machine:

```{r}
cl <- parallel::makeCluster(detectCores())
```

<div style="color: red">
Error in socketAccept(socket = socket, blocking = TRUE, open = "a+b")  : 
  all connections are in use
</div>

. . . 

```{r}
cl <- parallelly::makeClusterPSOCK(detectCores())
```

<div style="color: red">
Error: Cannot create 192 parallel PSOCK nodes. Each node needs
one connection, but there are only 124 connections left out of
the maximum 128 available on this R installation
</div>


## 3. `detectCores()` may return too many cores

Our code may not work on larger, modern machines.

Solution:

```{r}
parallelly::availableCores(constraints = "connections")
```

R option for maximum return value of `availableCores()`:  
`parallelly.availableCores.system`

Environment variable:  
`R_PARALLELLY_AVAILABLECORES_SYSTEM`, e. g.
`R_PARALLELLY_AVAILABLECORES_SYSTEM=120`


## 4. `detectCores()` does not give the number of "allowed" cores

### a) On a Personal Computer

* As developers, we don't know how many cores the user wishes to set aside
* Better to let user decide on how many cores to use for which purpose
* **Concurrent R sessions may result in > 100% of capacity!**
* This becomes very inefficient


## 4. `detectCores()` does not give the number of "allowed" cores {.smaller}

### b) On a Shared Computer

* Running all CPU cores by default will slow everything down for **all the users**
* User causing the problem may not be aware of it
* May be hard to troubleshoot

Solution:

```{r}
parallelly::availableCores()
```

User or system admin can limit the default number of CPU cores

Environment variable:  
`R_PARALLELLY_AVAILABLECORES_FALLBACK`, e. g. 
`R_PARALLELLY_AVAILABLECORES_FALLBACK=2`


## 4. `detectCores()` does not give the number of "allowed" cores {.smaller}

### c) A Shared Compute Cluster with Many Machines

* High-performance compute (HPC) clusters use a job scheduler
* Slurm example:  
  `sbatch --cpus-per-task=48 --mem=256G run_my_rscript.sh`
* `detectCores()` does not respect job scheduler settings

Solution:

```{r}
parallelly::availableCores()
```

* recognizes environment variables of common HPC schedulers, e. g.:
* Fujitsu Technical Computing Suite (PJM), Grid Engine (SGE), Load Sharing Facility (LSF), PBS/Torque, and Simple Linux Utility for Resource Management (Slurm)


## 4. `detectCores()` does not give the number of "allowed" cores {.smaller}

### d) Running R via CGroups in a Linux container

* croups = control groups: process gets allotted a certain number of cores
* Example: `docker run --cpuset-cpus=0-2,8` or `docker run --cpu=3.4 …`
* Problem: `detectCores()` returns CPUs of hardware, not of cgroup!
* Example: 96 cores, 8 cores per cgroup: 96 workers fight for resources of 8 cores; **very inefficient**
* **This can happen on Posit Cloud!**

Solution:

```{r}
parallelly::availableCores()
```


## Henrik Bengtsson's Recommendation

* Safest way: design R package to run sequentially by default
* Let the user decide on number of parallel workers to use

* Second-best alternative: Replace `detectCores()` with **`parallelly::availableCores()`**

For more details, see [https://parallelly.futureverse.org/](https://parallelly.futureverse.org/)


## Thanks!

<br>

### Youtube: StatistikinDD

### Twitter: @StatistikinDD

### github: fjodor

<br>
Presentation created using **quarto**, [https://quarto.org/](https://quarto.org/)
